Suppl. Figure 3. & 4. Mirobiome-KEGG correlations.

### 2022-01-07 reports/rmarkdown/manuscript/Suppl.Figure3-4.Rmd
Suppl. Figure 3 & 4

---
title: "`r params$project_title`"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
params:
  organisation: Karolinska Institutet
  pi_mail: prin.cipal@ki.se
  pi_name: Prin Cipal
  project_title: Broliden_5325
  redmine_issue: '#5325'
  requester_mail: kristina.broliden@ki.se
  requester_name: Kristina Broliden
  staff_mail: supp.ortstaff@nbis.se
  staff_name: Supp Ortstaff
  staff_web: https://nbis.se/about/staff/supp-ortstaff/
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hide",
                      message    = FALSE,
                      warning    = FALSE)
#setwd("/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/reports/rmarkdown")
```

#Load libraries and other scripts

```{r, echo = FALSE}
library(igraph)
library(rafalib)
library(fgsea)
library(scales)
remotes::install_github("czarnewski/niceRplots",upgrade = "always",force = T)
library(niceRplots)
library(tidyverse)
```

#Defining some variables for the analysis

```{r, echo = FALSE}
PATH <- "../../" #Path to the data
Set1 <- c("#4DAF4A", "#377EB8", "#E41A1C", "#FF7F00", "#984EA3", "#FFFF33", "#A65628", "#F781BF", "#999999")

#color palettes
pal <- c(Set1,RColorBrewer::brewer.pal(8,"Set2"),
         RColorBrewer::brewer.pal(9,"Pastel1"),RColorBrewer::brewer.pal(8,"Pastel2")) #color pallete for plots
heat_pal <- c("#000000", colorRampPalette(c("#000000","grey5","grey30","orange3","yellow","yellow","white"))(90))
cor_pal <- colorRampPalette(c("navy","white","firebrick"))(90)

#Graph construction  
fct <- .25 # FC threshold for differential expression
pvt <- 0.01 # Pvalue threshold for differential expression
min_pct <- .05 # minimun level of detected bacteria in each sample group

@@

# Loading data and metadata

```{r, echo = FALSE}
# Load meta data
meta <- read.csv(paste0(PATH,"/data/Clinical_visit_2_3_updatefeb22.csv"))
rownames(meta) <- as.character(meta$ID)

# Load datasets
dataset_names <- c("Tissue_RNAseq_V3_normalized",
                     "ASV_tissue_V3_normalized_batch_corrected",
                     "ASV_CVL_V3_normalized_batch_corrected",
                     "ASV_CVL_V2_normalized_NOT_batch_corrected")

#import datasets and return matrix with genus_lacto level resolution
# datasets <- map(dataset_names, 
#                 ~read.csv(paste0(PATH,"results/",.x, ".csv"),row.names = 1)) %>%
#   set_names(., dataset_names)

#fill all samples with 0s to make plotting easier
#x <- datasets[[2]]
# all_samples <- unique(unlist(lapply(datasets,colnames)))
# datasets_all_samples <- lapply(datasets, function(x) {
#   col <- setdiff(all_samples, colnames(x))
#   temp <- tibble(t = rownames(x))  %>% 
#      tibble::add_column(!!!set_names(as.list(rep(0, length(col))),nm=col)) %>% 
#     column_to_rownames(., var = "t") %>% bind_cols(x) 
#   return(temp)
# })

#import datasets and return matrix with genus_lacto level resolution
datasets <- lapply(dataset_names, function(x){
  x <- read.csv( paste0(PATH,"results/",x,".csv"),row.names = 1)
  return(as.matrix(x)) })
names(datasets) <- dataset_names

# Define what samples to use
samples_to_use <- intersect(colnames(datasets[["ASV_CVL_V3_normalized_batch_corrected"]]),
                            colnames(datasets[["Tissue_RNAseq_V3_normalized"]]))

#fill all samples with 0s to make plotting easier
x <- datasets[[1]]
datasets_all_samples <- lapply(datasets, function(x) {
  temp <- matrix(0,nrow = nrow(x),
                 ncol = nrow(meta),
                 dimnames = list(rownames(x),rownames(meta))
                 )
  temp[rownames(x),colnames(x)] <- x
  return(temp)
})
lapply(datasets_all_samples,dim)

# Clean up metadata
tissue_gr <- read.csv(paste0(PATH,"/data/210609_TissuegroupsJune.csv"))
metadata <- meta %>%
  left_join(tissue_gr, by="ID") %>%
  mutate(Luminal_names = ifelse(.$ModifiedM4 == "NonIners", "L. crispatus/jensenii",
                        ifelse(.$ModifiedM4 == "Iners", "L. iners",
                          ifelse(.$ModifiedM4 == "Gard", "Gardnerella",
                            ifelse(.$ModifiedM4 == "Prev", "High diverse", 
                                    "Other")))) ) %>%
  mutate(Luminal_gr = ifelse(.$ModifiedM4 == "NonIners", "L1",
                        ifelse(.$ModifiedM4 == "Iners", "L2", 
                          ifelse(.$ModifiedM4 == "Gard", "L3",
                            ifelse(.$ModifiedM4 == "Prev", "L4", 
                                    "L5")))) ) %>%
  mutate(Tissue_names = ifelse(.$TissuegroupsJune == "Crisp", "L. crispatus/jensenii",
                        ifelse(.$TissuegroupsJune == "Iners", "L. iners", 
                          ifelse(.$TissuegroupsJune == "Gard", "Gardnerella",
                            ifelse(.$TissuegroupsJune == "Prev", "High diverse", 
                                    "Other")))) ) %>%
  mutate(Tissue_gr = ifelse(.$TissuegroupsJune == "Crisp", "T1",
                        ifelse(.$TissuegroupsJune == "Iners", "T2", 
                          ifelse(.$TissuegroupsJune == "Gard", "T3",
                            ifelse(.$TissuegroupsJune == "Prev", "T4", 
                                    "T5")))) ) %>%
  select(ID:Yeast_v3, group_M9_CVLv3, Luminal_names, Tissue_names, Luminal_gr,Tissue_gr)

rownames(metadata) <- as.character(metadata$ID)
```

# Calculate QC metrics

\clearpage

```{r, echo = FALSE,fig.width=10,fig.height=5}
counts <- t(as.data.frame( lapply(datasets_all_samples,colSums) ))
counts[is.na(counts)] <- 0
detected_counts <- t(as.data.frame( lapply(datasets_all_samples,function(x) colSums( x > 0 )) ))
detected_counts[is.na(detected_counts)] <- 0
rownames(detected_counts) <- paste0(rownames(detected_counts),">0")

mypar(2,1,mar=c(2,10,2,1))
barlist( data = rbind( counts ),main = "total counts",
         genes = c(rownames(counts)), labels = sub("_norm.*","",rownames(counts)),
         clustering = rep(1,nrow(metadata)),
         col=ifelse( (colSums(counts == 0)==3)*1,"red", ifelse( (colSums(counts == 0)==2)*1,"orange3","grey" ) ),draw_mean_lines=F)
add_letter("a")

barlist( data = rbind( detected_counts),main = "detected features",
         genes = c(rownames(detected_counts)),labels = sub("_norm.*","",rownames(counts)),
         clustering = rep(1,nrow(metadata)),
         col=ifelse( (colSums(counts == 0)==3)*1,"red", ifelse( (colSums(counts == 0)==2)*1,"orange3","grey" ) ),draw_mean_lines=F)
add_letter("b")

sum((colSums(counts == 0)>=3)*1)
colnames(counts)[(colSums(counts == 0)==3)]
paste0(colnames(counts)[(colSums(counts == 0)==3)],collapse = ",")

metadata[colnames(counts)[(colSums(counts == 0)>=3)],]

datasets_all_samples <- lapply( datasets_all_samples , function(x) { x [ , samples_to_use] } )
metadata <- metadata[ samples_to_use , ]

detected_counts <- detected_counts[,samples_to_use]
barlist( data = rbind( detected_counts),main = "detected features",
         genes = c(rownames(detected_counts)),labels = sub("_norm.*","",rownames(counts)),
         clustering = rep(1,nrow(metadata)),
         col="grey",draw_mean_lines=F)
add_letter("b")

saveRDS(datasets_all_samples,"../../results/datasets_all_samples.RDS")
```

**Figure 1.** Comparative barplot for the **a)** total counts and **b)** number of non-zero detected features (genes / bacteria) for each of the sequencing datasets. Samples are ordered alfabetically according to the patient ID. Samples that are present in exactly two datasets are shown in orange (`r paste0(colnames(counts)[(colSums(counts == 0)==2)],collapse = ",")`). Samples that are present in exactly two datasets are shown in red (`r paste0(colnames(counts)[(colSums(counts == 0)==3)],collapse = ",")`).

```{r, echo = FALSE,fig.width=10,fig.height=9}
# all_microbiome <- rbind(datasets_all_samples[["ASV_tissue_V3_normalized_batch_corrected"]],
#                         datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]],
#                         datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]])
all_microbiome <- datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]]
# all_microbiome <- t( t(all_microbiome) / colSums(all_microbiome) )
all_microbiome <- all_microbiome[ rowSums(all_microbiome>0) >= 3 ,  ]
# all_microbiome <- all_microbiome + rnorm( length(all_microbiome) , mean = 0, sd = .00000001)
dim(all_microbiome)

zscore <- t(apply(all_microbiome,1,function(i){scale(i,T,T)}))
zscore[ zscore > 4 ] <- 4
zscore[ zscore < -4 ] <- -4
rownames(zscore) <- rownames(all_microbiome); colnames(zscore) <- colnames(all_microbiome)
mypar(mar=c(2,15,2,2))
top_var <- apply(all_microbiome,1,var)
top_var_t <- sort(top_var,decreasing = T)[100]
is_top_var <- top_var > top_var_t 
barplot(sort(top_var[is_top_var],decreasing = T)[30:1],horiz = T,las=1)

#Compute sample-wise correlations
cors <- cor(all_microbiome[,], method = "pearson")
adj <- (1-cors)/2
# pheatmap::pheatmap(adj,border_color = NA,color = colorRampPalette(c("grey90","grey70","grey70","grey70","orange3","firebrick"))(90) )

# cors <- cor(all_microbiome, method = "pearson")
# adj <- (1-cors)/2
# pheatmap::pheatmap(adj,border_color = NA,color = colorRampPalette(c("grey90","grey70","grey70","grey70","orange3","firebrick"))(90) )

#Graph construction  
k <- 12 # k for KNN
p <- 0.1 # prunning threshold of SNN 0.14

#Compute KNN graph from correlation matrix
knn <- RANN::nn2(adj,k = k)
nn <- matrix(0, nrow(knn$nn.idx), nrow(knn$nn.idx))
for(idx in 1:nrow(knn$nn.idx) ){  nn[idx,knn$nn.idx[idx,]] <- 1 }

#Computing SNN matrix
N <- matrix(1,ncol = ncol(nn),nrow = nrow(nn))
P <- t(nn) %*% nn
B <- N - nn
C <- t(B) %*% B
Q <- nrow(nn)*N - C
J <- P/Q
J[is.nan(J)] <- 0
J[J < p] <- 0
colnames(J) <- colnames(all_microbiome)
rownames(J) <- colnames(all_microbiome)

#Transforming SNN into an igraph object
g <- graph_from_adjacency_matrix(J,mode = "undirected",diag = F,weighted = T)
write.csv(J,"../../results/patient_SNN_graph.csv",row.names = T)

mypar(3,3,mar=c(0,0,2,0))

set.seed(1)
set.seed(1)
l <- layout_with_fr(g,niter=3000,start.temp=30)

cl <- igraph::cluster_louvain(g)
# cl <- igraph::walktrap.community(g)

metadata$Louvain_clusters <- cl$membership
metadata$joint_clustering <- metadata$Luminal_gr
metadata$joint_clustering <- factor(metadata$joint_clustering) 
set.seed(1)
U <- uwot::umap(adj,n_neighbors = 30)
h <- hclust(as.dist(adj),method = "ward.D2")
# metadata$joint_clustering <- cutree(h,k = 5)
write.csv2(metadata,"../../results/metadata_integration.csv",row.names = T)

plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(cl$membership)] , 
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,layout=l,
      vertex.size=10,main="FR embedding\nlouvain",
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )

plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(cl$membership)] , 
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,layout=U,
      vertex.size=10,main="umap",
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
empty_plot()
legend("topleft",title = "Patient groups",
       legend = levels(factor(cl$membership)),
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1.2,xpd=T)

plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(metadata$group_M9_CVLv3)] , 
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,layout=l,
      vertex.size=10,main="FR embedding",
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(metadata$group_M9_CVLv3)] , 
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,layout=U,
      vertex.size=10,main="umap",
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
empty_plot()
legend("topleft",title = "Patient groups\ngroup_M9_CVLv3",
       legend = levels(factor(metadata$group_M9_CVLv3)),
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1.2,xpd=T)
# title("Louvain clusters",cex.main=1,font.main=1)

plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(metadata$Luminal_gr)] , 
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,layout=l,
      vertex.size=10,main="FR embedding",
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(metadata$Luminal_gr)] , 
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,layout=U,
      vertex.size=10,main="umap",
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
empty_plot()
legend("topleft",title = "Patient groups\nLuminal_gr",
       legend = levels(factor(metadata$Luminal_gr)),
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1.2,xpd=T)
# title("Louvain clusters",cex.main=1,font.main=1)

@@

**Figure 1b.** Sample embedding of `r k`-SNN graph clustered using Louvain based on the normalized bacterial abundances across all microbiome datasets (including samples not present in all datasets).

```{r}
mypar(mar=c(3.2,15,2,2))

barlist( data = rbind( detected_counts),main = "detected features",
         genes = c(rownames(detected_counts)),labels = sub("_norm.*","",rownames(counts)),
         clustering = metadata$joint_clustering, 
         col=pal[factor(metadata$joint_clustering)],draw_mean_lines=F)
add_letter("b")

barlist( data = all_microbiome, main = "detected features",
         genes = c("Gardnerella",
                   "Lactobacillus iners",
                   "Lactobacillus jensenii",
                   "Lactobacillus crispatus/acidophilus",
                   "Prevotella",
                   "Atopobium"),
         clustering = metadata$joint_clustering,
         col=pal[factor(metadata$joint_clustering)], draw_mean_lines=F) # ,srt = 65
#legend(0,legend = levels(factor(metadata$Luminal_gr)))

# barlist( data = all_microbiome, main = "detected features",
#          genes = c("Gardnerella vaginalis",
#                    "Lactobacillus iners",
#                    "Lactobacillus jensenii",
#                    "Lactobacillus crispatus/acidophilus",
#                    "Prevotella amnii",
#                    "Prevotella timonensis",
#                    "Atopobium vaginae"),
#          clustering = metadata$joint_clustering,
#          col=pal,draw_mean_lines=F)
```

\clearpage

```{r, echo = FALSE,fig.width=5,fig.height=5}
x <- datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]]
xL <- rowsum(x , grepl("L.",rownames(x)) )[2,] / sum(grepl("L.",rownames(x)))
xM <- rowsum(x , grepl("Mobiluncus",rownames(x)) )[2,] / sum(grepl("Mobiluncus",rownames(x)))
xG <- rowsum(x , grepl("Gardnerella",rownames(x)) )[2,] / sum(grepl("Gardnerella",rownames(x)))

mL <- metadata$BV_Lactobacillus_v3 ; mL[is.na(mL)] <- 0
mM <- metadata$BV_Monbilicus_v3    ; mM[is.na(mM)] <- 0
mG <- metadata$BV_Vaginal_Garda_v3 ; mG[is.na(mG)] <- 0

mypar(3,1,mar=c(2,6,2,2))
barlist( data = rbind(  BV=mL ,
                        counts_16S=xL),
         main = "Lactobacillus",
         genes = c("BV","counts_16S"),
         clustering = metadata$joint_clustering, #srt = 60,
         draw_mean_lines=F,col = pal[factor(metadata$joint_clustering)])

barlist( data = rbind(  BV=mM ,
                        counts_16S=xM),
         main = "Monbilicus",
         genes = c("BV","counts_16S"), #srt = 60,
         clustering = metadata$joint_clustering,
         draw_mean_lines=F,col = pal[factor(metadata$joint_clustering)])

barlist( data = rbind(  BV=mG ,
                        counts_16S=xG),
         main = "Gardnerella",
         genes = c("BV","counts_16S"), #srt = 60,
         clustering = metadata$joint_clustering,
         draw_mean_lines=F,col = pal[factor(metadata$joint_clustering)])
```

\clearpage

```{r, eval=T,echo = FALSE,fig.width=10,fig.height=4}
myletters <- letters

mypar(1,2,mar=c(2,15,1,1))

for (m in c("ASV_tissue_V3_normalized_batch_corrected",
            "ASV_CVL_V3_normalized_batch_corrected",
            "ASV_CVL_V2_normalized_NOT_batch_corrected")){

  all_microbiome <- datasets_all_samples[[m]]

  datasets <- factor(metadata$joint_clustering)
  # datasets <- datasets[ colSums(all_microbiome)>0 ]
  # all_microbiome <- all_microbiome[,colSums(all_microbiome)>0]
  NN <- min(table(datasets))

  res <- data.frame( matrix(0,nrow = 1,ncol = 6) )
  for(i in levels(datasets)){
    for(j in rownames(all_microbiome) ){

      set.seed(1)
      a <- c(sample(all_microbiome[j,datasets == i],NN))
      set.seed(1)
      b <- sample(all_microbiome[j,datasets != i],NN)

      perc1 <- sum(all_microbiome[j,datasets == i]>0) / sum(datasets == i)
      perc2 <- sum(all_microbiome[j,datasets != i]>0) / sum(datasets != i)

      temp <- wilcox.test(x=a, y=b)

      fc <- log2( (mean(a)+1e-3) / (mean(b)+1e-3) )

      res <- rbind(res, setNames(c(j,i,fc,perc1,perc2,unlist(temp)[2] ),
                                 c("bacteria","cluster","fc","perc.1","perc.2","pvalue")) )
      colnames(res) <- c("bacteria","cluster","fc","perc.1","perc.2","pvalue")
    }
  }
  res <- res[-1,]
  res$pvalue <- as.numeric(res$pvalue)
  res$perc.1 <- as.numeric(res$perc.1)
  res$perc.2 <- as.numeric(res$perc.2)
  res$perc.diff <- res$perc.1 - res$perc.2

  res$fc <- as.numeric(res$fc)
  res$FDR <- p.adjust(res$pvalue)
  # res <- res[order(res$pvalue),]
  res <- res[abs(res$fc) > fct,]
  res <- res[abs(res$pvalue) < 0.05,]
  res <- res[ (res$perc.1 > min_pct) | (res$perc.2 > min_pct) ,]
  dim(res)

  ord <- factor(sapply(unique(as.character(res$bacteria)),function(x){getcluster(all_microbiome, x, datasets)}))

  plot_dots( all_microbiome , names(sort(ord)) , clustering = datasets,show_grid = T,cex.main=1,font.main=1,
            cex.row = .8,cex.col = .8,srt = 0,main = m)
  add_letter(myletters); myletters <- myletters[-1]

  barlist( all_microbiome , names(sort(ord)) , clustering = datasets,show_grid = T,cex.main=1,font.main=1,
            cex.axis = .8,srt = 0,main = m)
  add_letter(myletters); myletters <- myletters[-1]

}
```

**Figure 2.** Differential bacterial abundance across joint clustered samples and compared across microbiome datasets. The results are shown both as **a)** Dot plots and **b)** barplots Bacteria with log2FC above `r fct` and p-value below `r pvt` were considered significant and were sorted by the highest expression.

\clearpage

```{r, eval=T,echo = FALSE,fig.width=10,fig.height=9}
all_microbiome <- cbind(datasets_all_samples[["ASV_tissue_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]])
datasets <- factor(c( rep("tissue_v3",ncol(datasets_all_samples[["ASV_tissue_V3_normalized_batch_corrected"]])) , rep("CVL_v3",ncol(datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]])) , rep("CVL_v2",ncol(datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]])) ))
datasets <- datasets[ colSums(all_microbiome)>0 ]
all_microbiome <- all_microbiome[,colSums(all_microbiome)>0]

NN <- min(table(datasets))

res <- data.frame( matrix(0,nrow = 1,ncol = 6) )
for(i in levels(datasets)){
  for(j in rownames(all_microbiome) ){

    a <- all_microbiome[j,datasets == i]
    b <- all_microbiome[j,datasets != i]

    perc1 <- sum(all_microbiome[j,datasets == i]>0) / sum(datasets == i)
    perc2 <- sum(all_microbiome[j,datasets != i]>0) / sum(datasets != i)

    temp <- wilcox.test(x=a, y=b)

    fc <- log2( (mean(a)+1e-3) / (mean(b)+1e-3) )

    res <- rbind(res, setNames(c(j,i,fc,perc1,perc2,unlist(temp)[2] ),
                          c("bacteria","cluster","fc","perc.1","perc.2","pvalue")) )
    colnames(res) <- c("bacteria","cluster","fc","perc.1","perc.2","pvalue")
  }
}
res <- res[-1,]
res$pvalue <- as.numeric(res$pvalue)
res$perc.1 <- as.numeric(res$perc.1)
res$perc.2 <- as.numeric(res$perc.2)
res$perc.diff <- res$perc.1 - res$perc.2

res$fc <- as.numeric(res$fc)
res$FDR <- p.adjust(res$pvalue)
res <- res[order(res$pvalue),]
res <- res[abs(res$fc) > fct,]
res <- res[abs(res$pvalue) < pvt,]
res <- res[ (res$perc.1 > min_pct) | (res$perc.2 > min_pct) ,]
dim(res)

ord <- factor(sapply(unique(as.character(res$bacteria)),function(x){getcluster(all_microbiome, x, datasets)}))

mypar(1,2,mar=c(2,15,1,1))
plot_dots( all_microbiome , names(sort(ord)) , clustering = datasets,show_grid = T,cex.main=1,font.main=1,
          cex.row = .8,cex.col = .8,srt = 0)
add_letter("a")
barlist( all_microbiome , names(sort(ord)) , clustering = datasets,show_grid = T,cex.main=1,font.main=1,
          cex.axis = .8,srt = 0)
add_letter("b")
```

**Figure 2.** Differential bacterial abundance across microbiome datasets. The results are shown both as **a)** Dot plots and **b)** barplots Bacteria with log2FC above `r fct` and p-value below `r pvt` were considered significant and were sorted by the highest expression.

\clearpage

```{r, eval=T,echo = FALSE,fig.width=10,fig.height=9}
all_microbiome <- cbind(datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]])
datasets <- factor(c( rep("CVL_v3",ncol(datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]])) , rep("CVL_v2",ncol(datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]])) ))
datasets <- datasets[ colSums(all_microbiome)>0 ]
all_microbiome <- all_microbiome[,colSums(all_microbiome)>0]

NN <- min(table(datasets))

res <- data.frame( matrix(0,nrow = 1,ncol = 6) )
for(i in levels(datasets)){
  for(j in rownames(all_microbiome) ){

    a <- all_microbiome[j,datasets == i]
    b <- all_microbiome[j,datasets != i]

    perc1 <- sum(all_microbiome[j,datasets == i]>0) / sum(datasets == i)
    perc2 <- sum(all_microbiome[j,datasets != i]>0) / sum(datasets != i)

    temp <- wilcox.test(x=a, y=b)

    fc <- log2( (mean(a)+1e-3) / (mean(b)+1e-3) )

    res <- rbind(res, setNames(c(j,i,fc,perc1,perc2,unlist(temp)[2] ),
                          c("bacteria","cluster","fc","perc.1","perc.2","pvalue")) )
    colnames(res) <- c("bacteria","cluster","fc","perc.1","perc.2","pvalue")
  }
}
res <- res[-1,]
res$pvalue <- as.numeric(res$pvalue)
res$perc.1 <- as.numeric(res$perc.1)
res$perc.2 <- as.numeric(res$perc.2)
res$perc.diff <- res$perc.1 - res$perc.2

res$fc <- as.numeric(res$fc)
res$FDR <- p.adjust(res$pvalue)
res <- res[order(res$pvalue),]
res <- res[abs(res$fc) > fct,]
res <- res[abs(res$pvalue) < pvt,]
res <- res[ (res$perc.1 > min_pct) | (res$perc.2 > min_pct) ,]
dim(res)

>>>>

---
title: "Suppl. Figure 3. & 4. Mirobiome-KEGG correlations."
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes: 
- \usepackage{float}
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
##################
# LOAD LIBRARIES #
##################
#setwd("/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/reports/rmarkdown/manuscript")
suppressWarnings({suppressMessages({suppressPackageStartupMessages({
  library(tidyverse)
  library(openxlsx)
  library(fgsea)
  library(rafalib)
  library(Rcpp)
  #remotes::install_github("czarnewski/niceRplots",force=T)
  library(niceRplots)
})  })  })

#############
# LODA DATA #
#############
#Protein_V3 <- read_csv("../../../data/2020_04_20_LAMIQ_visit_3_singlets.csv") 
datasets_all_samples <- readRDS("../../../results/datasets_all_samples.RDS")
PATH <- "../../.."
SNN_bacteria <- read.csv("../../../results/bacteria_SNN_graph.csv",row.names = 1)
bac_communities <- read.xlsx("./Suppl.Tbl/Suppl.Tbl.01.xlsx", sheet = "Bacterial Communities")

@@

```{r echo=FALSE}
counts <- t(as.data.frame( lapply(datasets_all_samples,colSums) ))
counts[is.na(counts)] <- 0
```

```{r, warning=FALSE, echo = FALSE,fig.width=12,fig.height=13}
myletter <- letters
#reprod. <- FALSE
reprod. <- TRUE

ASV <- list(c("Luminal","ASV_CVL_V3_normalized_batch_corrected"),
            c("Tissue","ASV_tissue_V3_normalized_batch_corrected")
               )
# i <- ASV[[1]]
for(i in ASV){
  cat(paste0("Correlation between RNAseq and ",i[1], " 16S","\n"))

  # because hclust is non reproducible 
  if(reprod. == TRUE){
    NESes <- read.csv2(paste0("../../../results/",i[1],"_TRX_Bact_norm_enrichment_scores",".csv"), row.names = 1)
  }else{

    TRX <- datasets_all_samples[["Tissue_RNAseq_V3_normalized"]]
    bac_dataset <- datasets_all_samples[[i[2]]]
    samples_TRX_and_CVL3 <- colnames(counts) [ colSums(counts[c("Tissue_RNAseq_V3_normalized",i[2]),] > 0)==2 ]

    #filter TRX dataset
    TRX <- TRX [ , samples_TRX_and_CVL3 ]
    TRX <- TRX [ rowSums(TRX>0)>= 2 , ]
    # only top 5000 most variabe genes are included in the correlation
    top_vars_TRX <- names(sort(apply(TRX,1,var),decreasing = T)[1:5000] )
    TRX <- TRX[ top_vars_TRX , ]
    dim(TRX)

    #filter bac_dataset dataset
    bac_dataset <- bac_dataset [ , samples_TRX_and_CVL3 ]
    # keeping taxa that is present in at least two samples :
    bac_dataset <- bac_dataset [ rowSums(bac_dataset>0)>= 2 , ]
    top_vars_bacs <- names(sort(apply(bac_dataset,1,var),decreasing = T)[1:100] )
    dim(bac_dataset)

    cors <- cor( t(rbind(TRX) ) ,  t(rbind(bac_dataset) ) )

    bacteria_use <- colnames(cors)
    gmt <- gmtPathways(paste0(PATH, "/supplementary_files/c2.cp.kegg.v6.2.symbols.gmt.txt"))

    enrichments <- lapply(bacteria_use,gmt=gmt,cors=cors,function(x,gmt,cors){
      res <- fgsea(pathways = gmt,stats = sort(cors[,x],decreasing = T),nper=10000)
      return(res)
    })
    names(enrichments) <- bacteria_use

    pvalues <- lapply(enrichments,function(x) setNames(x$pval,x$pathway) )
    pvalues <- t(as.data.frame(pvalues))
    pvalues <- -log10( pvalues )
    top_pathways <- names(sort(apply(pvalues,1,median),T))[1:50]

    NESes <- lapply(enrichments,function(x) setNames(x$NES,x$pathway) )
    NESes <- t(as.data.frame(NESes))
    NESes[is.na(NESes)] <- 0
    # replace all NESes values with 0 if they don't have a significant p-value < 0.05:
    NESes[ pvalues < -log10(0.05) ] <- 0 
    # filter out bacterium and pathways with <10 significant NES scores 
    NESes <- NESes[rowSums(NESes!=0) >= 10 , colSums(NESes!=0) >= 10]

    set.seed(1)
    o_kegg <- hclust( as.dist( (1-cor(NESes))/2 ),"ward.D2")$order
    o_bacs <- hclust( as.dist( (1-cor(t(NESes) ))/2 ),"ward.D2")$order
    NESes <- NESes[o_bacs, o_kegg]

    bact <- gsub("\\.{1}", "/", rownames(NESes))
    bact <- gsub("\\/{2}", ". ", bact)
    rownames(NESes) <- bact

    #write.csv2(NESes, paste0("../../../results/",i[1],"_TRX_Bact_norm_enrichment_scores",".csv"))

  terms <- gsub("_", " ", colnames(NESes))
  terms <- gsub("(GO |KEGG )", "", terms, perl = TRUE)
  colnames(NESes) <- terms

  par(mfrow=c(1,1),mar=c(12,0,2,4)) #b,l,t,r
  image( t(NESes[nrow(NESes):1,]),col=colorRampPalette(c("navy","grey95","firebrick"))(91),breaks=seq(-3,3,length.out = 92),
         axes=F,border=NA,main="tissue RNAseq KEGG pathways",xlab="",ylab=i,line=.4,cex.main=1,font.main=1)
  text(  par("usr")[c(4)] , seq(1,0,length.out = nrow(NESes)),
        labels = rownames(NESes), srt = 0, adj = c(0,.5), xpd = TRUE, cex=.4)
  text( seq(0,1,length.out = ncol(NESes)) , par("usr")[c(1)],
        labels = colnames(NESes), srt = 90, adj = c(1,.5), xpd = TRUE, cex=.4)

@@

**Figure 9.** Heatmaps showing the functional association of microbiome datasets with the expression profiles in the RNAseq dataset. Briefly, bacterial abundances from each dataset were correlated with the gene expression of the top 5000 highly variable genes from the RNAseq dataset, generating a correlation matrix between bacteria and genes. Then for each bacteria, we rank genes based on their correlation to that bacteria and perform gene set enrichment anlaysis (GSEA) using the KEGG gene annotation database. This, in turn, will result in a matrix associating every bacteria with every KEGG process in the tissue. The heatmap shows the normalized enrichment score (NES). Only enrichments with pvalue below 0.05 are shown.Only bacterium and pathways with at least 10 significant NES scores were included in the heatmap.

```{r eval=FALSE, include=FALSE}
g <- graph_from_adjacency_matrix(as.matrix(SNN_bacteria), mode = "undirected",diag = F,weighted = T)
cl <- igraph::cluster_louvain(g)
cors <- cor(t(datasets_all_samples[[i]][rowSums(datasets_all_samples[[i]] > 0)>=10,]))

module_means <- bac_communities %>%
  left_join(as_tibble(datasets_all_samples[[i]], rownames = "Taxa"), by="Taxa")  %>%
  group_by(bacterial.communities) %>%
  summarise(across(where(is.numeric), ~sum(.x))) %>%
  mutate(across(where(is.numeric), ~.x/c(table(bac_communities$bacterial.communities))) ) %>%
  select(-bacterial.communities) %>%
  #column_to_rownames(var="bacterial.communities") %>%
  as.matrix() %>%
  `rownames<-`(unique( bac_communities$bacterial.communities)) 
bact_df <- module_means

all_microbiome <- cbind(datasets_all_samples[["ASV_tissue_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]])
#cors <- cor( t(rbind(TRX) ) ,  t(rbind(bac_dataset) ) )

module_means_per_dataset <- lapply(datasets_all_samples[[i]],function(x){
  return( rowsum( x[bac_communities$Taxa,] , group = bac_communities$bacterial.communities ) / c(table(bac_communities$bacterial.communities)) )
})

module_means <- bac_communities %>%
  left_join(as_tibble(datasets_all_samples[[i]], rownames = "Taxa"), by="Taxa")  %>%
  group_by(bacterial.communities) %>%
  summarise(across(where(is.numeric), ~sum(.x))) %>%
  mutate(across(where(is.numeric), ~.x/c(table(bac_communities$bacterial.communities))) )

b  <-  datasets_all_samples[[i]][bac_communities$Taxa,]
rowsum(datasets_all_samples[[i]][bac_communities$Taxa,] , group = bac_communities$bacterial.communities )
```

```{r eval=FALSE, fig.height=8, fig.width=10, include=FALSE}
myletter <- letters
